COFFEE: an Optimizing Compiler\\for Finite Element Local Assembly

### 2014-04-05 paper/bare_conf.tex
Strip trailing whitespace
Some rewording of background section

\begin{algorithm}[t]
\small
\KwIn{stiffness matrix A (initialized to 0),\\~~~~~~~~~element coordinates in the domain (array),\\~~~~~~~~~(optional) fields (e.g. velocity, array)}
\KwOut{stiffness matrix A (computed)}
\nlset{1}~~- Jacobian computation, using element coordinates \\
~~- Declaration of constant data, among which basis function\\
~~- Computation of the stiffness matrix
\caption{General structure of a local assembly kernel generated by Firedrake.}
\label{code:general-structure}
\end{algorithm}

@@

Local assembly consists of evaluating so called element stiffness matrix (``stiffness matrix'', in the following) and element stiffness vector; in this work, we focus on computation of stiffness matrices, which is usually the costly part of the process. A stiffness matrix can be intuitively thought as an approximated representation of the PDE solution in a specific cell (or element) of the discretized domain. Numerical integration algorithms are widely-used in assembly codes to evaluate stiffness matrices~\cite{quadrature1, fluidity_manual_v4}.

Given a mathematical description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl}, Firedrake generates C-code kernels implementing assembly using a numerical integration algorithm. It then triggers compilation of such kernels using an available vendor compiler, and eventually manages parallel execution over the mesh. The subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code. The code transformations described next are also generalizable to non-Firedrake assembly kernels, provided that numerical integration is used.

The general structure of a Firedrake-generated kernel is shown in Figure~\ref{code:general-structure}. Input are a pointer to a 0-valued bi-dimensional array, which will be used to store the stiffness matrix; a pointer to the element's coordinates in the discretized domain; one or more pointers to so called fields, for instance indicating value of velocity or pressure in the element. The output is the evaluated element stiffness matrix. A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse; 2) definition of some constant bi-dimensional arrays, called ``basis functions'' (or derivatives of basis functions) in finite element terminology, initialized to problem-specific double-precision floating point values; 3) evaluation of the stiffness matrix in an affine loop nest. All elements in the discretized domain share the same basis function arrays, so they are declared as global read-only arrays (i.e. using \texttt{static const} in C language). Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to the various kernel's objects.

\begin{table}[h]
\begin{center}
\begin{tabulary}{1.0\columnwidth}{|C|C|C|}
\hline
Object name & Type & Variable name(s) \\\hline\hline
Determinant of Jacobian matrix & double & det  \\ \hline
Inverse of Jacobian matrix & double[] & K \\ \hline
Coordinates & double** & coords\\ \hline
Fields & double** & w \\ \hline
Integration points & double[] & W \\ \hline
Basis functions (and derivatives) & double[][] & X, Y, X1, ... \\ \hline
Stiffness matrix & double[][] & A\\ \hline
\end{tabulary}
\end{center}
\caption{Type and variable names used in the various code snippets to identify local assembly objects.}
\label{table:map-name-letters}
\end{table}

The actual complexity of a local assembly kernel depends on the finite element problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products. An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other scenarios, for instance when solving a non-linear problem like Burgers as in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the stiffness matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse. Note that in addition to a larger number of operations to compute the stiffness matrix, the Burgers case shows a register pressure higher than that in Helmholtz. Despite assembly kernels being problem-dependent, meaning that the space of codes that Firedrake can generate is infinite, it is still possible to identify common domain-specific traits that can be exploited for effective code transformations and SIMD vectorization.

Some peculiarities of assembly kernels are: 1) memory accesses along the three loop dimensions are always stride-1; 2) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g. the $F$ values in Burgers) and introduction of temporary arrays; 3) the $j$ and $k$ loops iterate over the same iteration space; 4) most of the sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$). In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.

>>>>

\begin{algorithm}[t]
\small
\KwIn{coordinate field (array), coefficient fields (e.g. velocity, array)}
\KwOut{element matrix A (2D array)}
\nlset{1}Compute Jacobian from coordinates \\
Declare of constant data: basis functions and derivatives\\
Compute element matrix using numerical quadrature
\caption{General structure of a local assembly kernel generated by Firedrake.}
\label{code:general-structure}
\end{algorithm}

@@

% FIXME: stiffness matrix is only used for certain types of elliptic PDEs
% FIXME: the fluidity manual isn't a good source for this, cite e.g. Brenner and Scott

Local assembly is the computation of local contributions of a specific cell of the discretized domain to the global sparse matrix or right-hand side vector of the linear system. The process consists of numerically evaluating an integral to produce a small dense element matrix or an element vector~\cite{quadrature1, fluidity_manual_v4}. In this work we focus on local matrices, which are usually more costly to compute.

% FIXME: you can't attribute the kernel code generation to Firedrake: mention FFC

Given a mathematical description of the input problem, expressed through the domain-specific Unified Form Language~\cite{ufl}, Firedrake generates C-code kernels implementing assembly using a numerical integration algorithm. These kernels are compiled at runtime using a vendor compiler, and Firedrake manages the parallel execution over the mesh. The subject of this paper is to enhance this execution model by adding an optimization stage prior to the generation of C code. The code transformations described next are also generalizable to non-Firedrake assembly kernels, provided that numerical integration is used.

The general structure of a Firedrake-generated kernel is shown in Figure~\ref{code:general-structure}. Inputs are a pointer to a zero initialized two dimensional array used to store the element matrix, a pointer to the element's coordinates in the discretized domain, and pointers to data of coefficient fields, for instance indicating value of velocity or pressure in the element. The output is the evaluated element matrix. A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse; 2) definition of some constant two dimensional arrays for the finite element basis functions and their derivatives (if needed), initialized to problem-specific double-precision floating point values; 3) evaluation of the element matrix in an affine loop nest. All elements in the discretized domain share the same basis function arrays, so they are declared as global read-only arrays (using \texttt{static const} in C). Table~\ref{table:map-name-letters} shows the variable names we will use in the upcoming code snippets to refer to various objects in the kernel.

\begin{table}[h]
\begin{center}
\begin{tabulary}{1.0\columnwidth}{|C|C|C|}
\hline
Object name & Type & Variable name(s) \\\hline\hline
Determinant of Jacobian matrix & double & det  \\ \hline
Inverse of Jacobian matrix & double[] & K \\ \hline
Coordinates & double** & coords\\ \hline
Fields & double** & w \\ \hline
Integration points & double[] & W \\ \hline
Basis functions (and derivatives) & double[][] & X, Y, X1, ... \\ \hline
Element matrix & double[][] & A\\ \hline
\end{tabulary}
\end{center}
\caption{Type and variable names used in the various code snippets to identify local assembly objects.}
\label{table:map-name-letters}
\end{table}

% FIXME: You haven't really defined basis functions, so what is a Lagrange basis function?

The actual complexity of a local assembly kernel depends on the finite element problem being solved. In simpler cases, the loop nest is perfect: it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products. An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other scenarios, for instance when solving a non-linear problem like Burgers as shown in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse. Note that in addition to a larger number of operations to compute the element matrix, the Burgers case shows a register pressure higher than that in Helmholtz. Despite assembly kernels being problem-dependent, meaning that the space of codes that Firedrake can generate is infinite, it is still possible to identify common domain-specific traits that can be exploited for effective code transformations and SIMD vectorization.

Some characteristics of assembly kernels are: 1) memory accesses along the three loop dimensions are always stride-1; 2) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g. the $F$ values in Burgers) and introduction of temporary arrays; 3) the $j$ and $k$ loops iterate over the same iteration space; 4) most of the sub-expressions on the right hand side of the element matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$). In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.
